Network based multifactorial modelling microRNA-target interaction.

### 2019-07-26 article.Rmd
finished results section

@@

(#TODO caption needs more text, every color should be explained)

@@

(#TODO about formulas, 1) eq:3 F/Fmax can be removed, 2) in formulas eq:1 and eq:2, C_m or Eff_gi is not intuitive, is there consensus terms about effect or nth gene? 3) in formula eq:4, instead of, 4) in general, `*` sign should be replaced with `\times` and apostrophe should be replaced with `\prime` )

@@

### Multifactorial networks
(#TODO title should be more informative)
It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we have improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We have explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. But these factors can be handled optionally and differently.

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2. When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions have observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation. On the other hand, It was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors.This also shows that the perturbation-initiation element in the system is important parameter. So, we also have developed an approach that is found perturbation efficiencies of elements in dataset. **(see details in Supplementary File(fig1_2app.Rmd)**. When we applied the method on the minimal dataset, while Gene4 is the most efficient element in terms of number of perturbed elements, it have observed that Mir2 is most potent element in the mean of expression change. In our approach, the elements behave differently based on type (competing or miRNA). When a competing element was changed, whole elements out of miRNAs are re-regulated. But, if a miRNA was perturbed, whole competing elements are re-regulated and miRNA preserves expression at after perturbation.

The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. We have tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Firstly, we have observed the activities in the iterations and determined behaviors of the network after simulation. As expected, we have observed that the big dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviors and complexity to gain steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibit competing behaviors in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, some have changed slightly. So, we have compared two datasets; the tumor gene expression and simulation results. When we analyzed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene and miRNA expression values at initial conditions. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure.

The network based approaches have been developed in previous studies. Figliuzzi et al. have tried to explain the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs. They have pointed out that the larger miRNA number can cause to evident crosstalk between ceRNAs. It has been thought that miRNAs and their targets interactions depend on rates of transcription, degradation, binding and unbinding in the network based kinetic model which has been developed by Nitzan et al. Nitzan and collaborators have used the high throughput experimental dataset [@helwak_mapping_2013] for miRNA:target pair that occur high free energy and microarray dataset of a miR-92a depletion for expression values. They have been demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions change to depend on distance between ceRNAs, and topological features of network [@nitzan_interactions_2014]. List et al. have developed an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets [@markus_list_sponge_2017]. Their approach could be useful to analysis interacted genes through miRNAs. When it has been thought that a miRNA can exhibit strong functionality to a target but may not against an other, common miRNA based approach can be inconvenient to understand regulations of ceRNA interactions.

In our approach, we have not taken into account transcription, degradation or binding rates of elements in network. Because, although it is known as the miRNAs are highly stable, the transcription and degradation rates of miRNAs change to depend on according to cellular conditions [@ruegger_microrna_2012]. According to our search, there is no dataset that was includes to degradation and transcription rates on specific cellular conditions for each miRNA. So, we preferred to accept that degradation and transcription rates of each miRNAs are equal to each other, instead of determining a constant value for all miRNAs in the system (The same applies to mRNA.). On the other hand other regulation parameters such as gene-gene interactions and transcription factors are ignored but the network structure can promote to integration of other regulation elements. In the future, with development in experimental techniques about features of miRNAs and their targets, more consistent and useful results can be obtained from our approach. Additionally, this can provide to contribution to prediction of abnormal regulations and pathways with the studies that will be developed.

(#TODO BRCA results go here) ABCC1 gene, *Multidrug resistance-associated protein 1*, is one of the most significant factor to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug]

>>>>

[#TODO]: <> "there are many et. al and et. al. , with or without dot(.) please make sure the formatting is correct and consistent"

@@

[#TODO]: <> "caption needs more text, every color should be explained"

@@

[#TODO]: <> "about formulas, 1) eq:3 F/Fmax can be removed, 2) in formulas eq:1 and eq:2, C_m or Eff_gi is not intuitive, is there consensus terms about effect or nth gene? 3) in formula eq:4, instead of, 4) in general, `*` sign should be replaced with `\times` and apostrophe should be replaced with `\prime` )"

@@

### Multifactorial networks

[#TODO]: <> "title should be more informative"

Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in  Figure \autoref{fig1} is recalculated with additional factors in effect [][#TODO] with same trigger, two fold increase in Gene2 (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed (compare miRNA distribution values between Figure \autoref{fig1}a and Figure \autoref{fig3}a). When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. [][#TODO2]

[#TODO]: <> "although data is visible at Figure2, since it's a small  table,  we might add the table showing expression levels, seed type, binding energy for M1, M2 and genes so that reader can see which miRNA  has low affinity and understand Figure 3 better"
[#TODO2]: <> "we should compare and contrast Figure 1a and figure 3a, the reader should see how affinity effects distrbution G4:M2, and how efficiency is effected without change in affnity M1:G3"

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), the changes of other gene expressions have observed more prominent [][#TODO]. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary File(fig1_2app.Rmd)**. When we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes. [][#TODO2]

[#TODO]: <> "didnt we perform upregulation of Gene2, why did we say Gene4 here?"
[#TODO2]: <> "Removed these sentences because it was somewhat confusing: In our approach, the elements behave differently based on type (competing or miRNA). When a competing element was changed, whole elements out of miRNAs are re-regulated. But, if a miRNA was perturbed, whole competing elements are re-regulated and miRNA preserves expression at after perturbation."

[][#TODO] We used our model to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. Simulations show that change in expression level of single gene has potential to effect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

[#TODO]: <> "removed this section, didnt understand, confusing: So, we have compared two datasets; the tumor gene expression and simulation results. When we analyzed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene and miRNA expression values at initial conditions. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure."

[#TODO]: <> "removed this section, was confusing: The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. "

Network based approaches for analyzing miRNA:target interactions have been developed in earlier studies. An initial  attempt demonstrated the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs [@figliuzzi_micrornas_2013]. Next, a network based kinetic model integrating miRNA and target rates of transcription, degradation, binding and unbinding was developed [@nitzan_interactions_2014] using high throughput experimental dataset about miR-92a depletion [@helwak_mapping_2013]. In this study, it was demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions are effected depending on distance between ceRNAs or topological features of network [@nitzan_interactions_2014]. More recently, an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets was developed [@markus_list_sponge_2017] which was effective in analyzing genes through miRNAs. Based on the observations that a miRNA can exhibit strong functionality to a target but may not against an other, the authors have concluded that common miRNA based approach may not be suitable for understanding regulations of ceRNA interactions.

[#TODO]: <> "we can talk about BRCA results here: ABCC1 gene, *Multidrug resistance-associated protein 1*, is one of the most significant factor to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug]"

In our approach, we have not taken into account transcription, degradation or binding rates of elements in network. Because, although it is known as the miRNAs are highly stable, the transcription and degradation rates of miRNAs change depending on cellular conditions [@ruegger_microrna_2012]. However, due to lack of datasets including degradation and transcription rates of miRNAs at specific cellular conditions, such factors are neglected in our model. Additionally, other regulation parameters such as gene-gene interactions and transcription factors are ignored but the network structure is flexible and can integrate additional regulation elements as they become available. In the future, with developments in experimental techniques about miRNAs:target interaction dynamica, more accurate results can be obtained from our model resulting in better understanding and predictions of abnormal regulations and pathways underlying diseases or conditions.
